A Horrocks-type theorem for even orthogonal $K_2$

### 2019-10-02 dl-paper.tex
Minor

For shortness we denote the subgroup $\overline{\St}(\Phi, A, M)$ by $G_M^0$.
Notice that $G_M^0$ can be considered as a subgroup of $\St(\Phi, R)$. Consider the following commutative diagram.
\[ \xymatrix{ G_M^0 \ar@{^{(}->}[r]^{i_+} \ar@{^{(}->}[d]^{i_-} & \overline{\St}(\Phi, A[X], M[X]) \ar^{j_+^M}[d] \ar@{^{(}->}[r] & \St(\Phi, A[X]) \ar[dl]^{j_+} \\
              \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, R). &  } \] 
We denote by $\overline{G}^{\geq 0}_M$ the image of the map $j_+^M$.               
Denote by $\overline{V}_T$ the quotient of the set of triples 
\[V_T = \overline{G}_M^{\geq 0} \times \St(\Phi, A[X\inv]) \times (1+M)^\times\] by the equivalence relation given by $(p \gamma, h, u)_T \sim (p, \gamma h, u)_T$, $\gamma \in G^0_M$. We denote the image of $(p, h, u)_T$ in $\overline{V}_T$ by $[p, h, u]$.              
$\overline{V}_T$ is precisely the set upon which M.~Tulenbaev in~\cite[Proposition~4.3]{Tu83} constructs an action of $\St(\Phi, B)$.

@@

This action satisfies the following additional properties.
\begin{enumerate}
 \item For any $h_1 \in \St(\Phi, A[X\inv])$ holds $j_B^-(h_1) \cdot [1, h, u] = [1, h_1 h, u]$, where $j_B^-$ denotes the map $\St(\Phi, A[X\inv]) \to \St(\Phi, B)$ (we identify $\overline{V}$ with $\overline{V}_T$ using the isomorphism~\eqref{eq:VVT}).
 \item If we consider $\St(\Phi, A[X, X\inv])$ as a group with the left multiplication action of $\St(\Phi, B)$ then the map $\overline{V} \to \St(\Phi, A[X, X\inv])$ given by $(g, [h], u) \mapsto g$ is a map of $\St(\Phi, B)$-groups.
\end{enumerate}
\end{prop}
\begin{proof} By~\cref{superfluous-relations} and Lemmas~\ref{R3_leq0_1}--\ref{R2_0_1} the action of $\St^{\leq 1}(\Phi, B)$ on $\overline{V}$ given by~\eqref{T_1} and~\eqref{T_leq0} is well-defined. On the other hand, the group $\St^{\leq 1}(\Phi, B)$ is isomorphic to $\St(\Phi, B)$ by~\cref{lemma33} (set $t = X^{-1}$).

@@

From the second property we also immediately obtain the following group factorization.
\begin{corollary} For $\Phi$ as above one has $\E(\Phi, A[X\inv] + M[X]) = \E(\Phi, A[X], M[X]) \cdot \E(\Phi, A[X\inv]).$ \end{corollary}

@@

\begin{lemma}\label{lem:action} For any $[p, h, u]\in \overline{V}_T$ and any $p_1 \in \overline{G}^{\geq 0}_M$ holds $p_1 \cdot [p, h, u] = [p_1p, h, u]$. \end{lemma}
\begin{proof} First of all, notice that the assertion of the lemma can be verified directly in the case $p_1 = z_\alpha(f, a)$ for all $a\in A$, $f\in M[X]$. Using~\cref{SR:additivity} and the fact that $x_\beta(X^2f) \in P_\alpha(0)$ for all $\beta \in \Phi$ it is not hard to show that the assertion also holds for $p_1 = z_\alpha(X^2f, aX^{-1})$.

@@

Suppose that the assertion holds for all $p_1 = z_\alpha(f, \xi)$ for which $\xi$ has degree $\leq n$.
Substituting $s := fX$, $\eta := X^{-1}$, $\xi := X\xi$ in~\eqref{Z5} we obtain the following equality in $\St(\Phi, B)$:
\begin{multline*}
 z_{\alpha+\beta}(f, X\xi) = x_\alpha(\epsilon Xf) \cdot x_{-\beta}(-faX^2) \cdot x_{\beta}(af) \cdot x_{\alpha+\beta}(f) \cdot \\ \cdot z_\alpha(-\epsilon Xf, -\epsilon \xi) \cdot x_{-\alpha}(-\epsilon \xi^2fX) \cdot x_{-\alpha-\beta}(- f \xi^2X^2) \cdot z_{-\beta}(X^2f\xi, -X^{-1}),\text{ where $\epsilon = N_{\alpha, \beta}$.}
\end{multline*}
From the first part of the proof and inductive assumption we obtain that the assertion of the lemma holds for all the factors in the right-hand side and, therefore, also holds  for $p_1 = z_\alpha(f, X\xi)$. It is easy to deduce from this that the assertion also holds for $p_1 = z_\alpha(f, X\xi + a) = x_\alpha(-a) \cdot z_\alpha(f, X\xi) \cdot x_\alpha(a)$. 
\end{proof}

\begin{rem}
Although we will not need this for our main result, it can be noted that $\overline{V}$ is a left $\St(\Phi, B)$-torsor, i.\,e. the action of $\St(\Phi, B)$ on $\overline{V}$ is both transitive and faithful. The faithfulness follows from the second property of~\cref{prop43} and our assumption that $j_R$ is injective.
The transitivity follows from~\cref{lem:action}, the first property of~\cref{prop43} and the following computation, which we leave as an exercise to the reader:
\begin{equation*} \langle 1, m \rangle^{-1} \langle X^{-1}, mX \rangle \cdot [1, 1, 1] = [1, 1, 1+m]. \end{equation*} \end{rem}

\begin{externaltheorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{externaltheorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g \in G_M^0$ and consequently $g = 1$.

@@

 Set $h_+ = g_+g_0^{-1}$ and $h_- = g_-g_0^{-1}$.
 It is clear that \[ h_+ \in \overline{\St}(\Phi, A[X], M[X]), h_- \in \overline{\St}(\Phi, A[X\inv], M[X\inv])\text{ and that }j_-(h_-) = j_+(h_+).\]
 From~\cref{prop43} and~\cref{lem:action} we obtain that
 \[ [j_+(h_+)^{-1}, h_-, 1] = j_+(h_+)^{-1} \cdot [1, h_-, 1] = j_+(h_+)^{-1} j_B^-(h_-) \cdot [1, 1, 1] = [1, 1, 1]. \]
 Thus, $h_-$ (and hence $h_+$) come from some element of $G_M^0$, which implies that $g_+$ and $g_-$ are images of some element of $\St(\Phi, A)$.
 \end{proof}

\begin{proof}[Proof of~\cref{thm:main}]
 Notice that the assertions of the theorem for $\KO_2(2\ell, -)$ and $\K_2(\rD_\ell, -)$ follows from the assertion for $\St(\rD_\ell, -)$.
 In turn, the latter assertion follows from~\cref{thm41} and~\cref{thm:P1glueing} in the special case when $A$ is a local ring.

>>>>

For shortness we denote the subgroup $\overline{\St}(\Phi, A, M)$ by $G_M^0$.
Notice that $G_M^0$ can be considered as a subgroup of $\St(\Phi, R)$. Consider the following commutative diagram.
\[ \xymatrix{ G_M^0 \ar@{^{(}->}[r]^{i_+} \ar@{^{(}->}[d]^{i_-} & \overline{\St}(\Phi, A[X], M[X]) \ar^{j_+^M}[d] \ar@{^{(}->}[r] & \St(\Phi, A[X]) \ar[dl]^{j_+} \\
              \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, R). &  } \] 
We denote by $\overline{G}^{\geq 0}_M$ the image of the map $j_+^M$.               
Denote by $\overline{V}_T$ the quotient of the set of triples 
\[V_T = \overline{G}_M^{\geq 0} \times \St(\Phi, A[X\inv]) \times (1+M)^\times\] by the equivalence relation given by $(p \cdot j_+(i_+(\gamma)), h, u)_T \sim (p, i_-(\gamma) \cdot h, u)_T$, $\gamma \in G^0_M$. We denote the image of $(p, h, u)_T$ in $\overline{V}_T$ by $[p, h, u]$.              
$\overline{V}_T$ is precisely the set upon which M.~Tulenbaev in~\cite[Proposition~4.3]{Tu83} constructs an action of $\St(\Phi, B)$.

@@

This action satisfies the following additional properties.
\begin{enumerate}
 \item For any $h_1 \in \St(\Phi, A[X\inv])$ holds $j_B^-(h_1) \cdot [1, h, u] = [1, h_1 h, u]$, where $j_B^-$ denotes the map $\St(\Phi, A[X\inv]) \to \St(\Phi, B)$ (we identify $\overline{V}$ with $\overline{V}_T$ using the isomorphism~\eqref{eq:VVT}).
 \item If we consider $\St(\Phi, A[X, X\inv])$ as a set with the left multiplication action of $\St(\Phi, B)$ then the map $\overline{V} \to \St(\Phi, A[X, X\inv])$ given by $(g, [h], u) \mapsto g$ is a map of $\St(\Phi, B)$-sets.
\end{enumerate}
\end{prop}
\begin{proof} By~\cref{superfluous-relations} and Lemmas~\ref{R3_leq0_1}--\ref{R2_0_1} the action of $\St^{\leq 1}(\Phi, B)$ on $\overline{V}$ given by~\eqref{T_1} and~\eqref{T_leq0} is well-defined. On the other hand, the group $\St^{\leq 1}(\Phi, B)$ is isomorphic to $\St(\Phi, B)$ by~\cref{lemma33} (set $t = X^{-1}$).

@@

For a pair $(R, I)$ denote by $\E(\Phi, R, I)$ the {\it relative elementary subgroup} of $G(\Phi, R)$, i.\,e.
the image of the relative Steinberg group $\overline{\St}(\Phi, R, I)$ under the map $\pi \colon \St(\Phi, R) \to G(\Phi, R)$.
From the second property of the above proposition we immediately obtain the following group factorization.
\begin{corollary} For $\Phi$ as in the statement of~\cref{lemma33} one has
\[\E(\Phi, A[X\inv] + M[X]) = \E(\Phi, A[X], M[X]) \cdot \E(\Phi, A[X\inv]).\] \end{corollary}

@@

\begin{lemma}\label{lem:action} For any $[p, h, u]\in \overline{V}_T$ and any $p_1 \in \overline{G}^{\geq 0}_M$ holds $p_1 \cdot [p, h, u] = [p_1p, h, u]$. \end{lemma}
\begin{proof} First of all, notice that the assertion of the lemma can be verified directly in the case $p_1 = z_\alpha(f, a)$ for all $a\in A$, $f\in M[X]$. Using~\eqref{eq:Smult} and the fact that $x_\beta(X^2f) \in P_\alpha(0)$ for all $\beta \in \Phi$ it is not hard to show that the assertion also holds for $p_1 = z_\alpha(X^2f, aX^{-1})$.

@@

Suppose that the assertion holds for all $p_1 = z_\alpha(f, \xi)$ for which $\xi$ has degree $\leq n$.
Substituting $s := fX$, $\eta := X^{-1}$, $\xi := X\xi$ into~\eqref{Z5} we obtain the following equality in $\St(\Phi, B)$:
\begin{multline*}
 z_{\alpha+\beta}(f, X\xi) = x_\alpha(\epsilon Xf) \cdot x_{-\beta}(-X^2 f\xi ) \cdot x_{\beta}(\xi f) \cdot x_{\alpha+\beta}(f) \cdot \\ \cdot z_\alpha(-\epsilon Xf, -\epsilon \xi) \cdot x_{-\alpha}(-\epsilon X\xi^2f) \cdot x_{-\alpha-\beta}(- X^2f \xi^2) \cdot z_{-\beta}(X^2f\xi, -X^{-1}),\text{ where $\epsilon = N_{\alpha, \beta}$.}
\end{multline*}
From the first part of the proof and inductive assumption we obtain that the assertion of the lemma holds for all the factors in the right-hand side and, therefore, also holds  for $p_1 = z_\alpha(f, X\xi)$. It is easy to deduce from this that the assertion also holds for $p_1 = z_\alpha(f, X\xi + a) = x_\alpha(-a) \cdot z_\alpha(f, X\xi) \cdot x_\alpha(a)$. 
\end{proof}

\begin{rem}
Although we will not need this for our main result, it can be noted that $\overline{V}$ is a left $\St(\Phi, B)$-torsor, i.\,e. the action of $\St(\Phi, B)$ on $\overline{V}$ is both transitive and faithful. The faithfulness follows from the second property of~\cref{prop43} and our assumption that $j_R$ is injective.
The transitivity follows from~\cref{lem:action}, the first property of~\cref{prop43} and the following calculation, which we leave as an exercise to the reader:
\begin{equation*} \langle 1, m \rangle^{-1} \langle X^{-1}, mX \rangle \cdot [1, 1, 1] = [1, 1, 1+m]. \end{equation*} \end{rem}

\begin{comment}
\begin{equation} \label{T_2} t_\alpha(aX^{-1}) (g, h, u) = \left[ x_\alpha(aX^{-1})\cdot g \cdot x_\alpha\left(-\tfrac{aX^{-1}}{1 + am}\right) \{ X, 1+am \} ,\ x_\alpha\left(\tfrac{aX^{-1}}{1 + am}\right) h ,\ u \cdot (1 + am)\right].\end{equation}

\begin{multline*}
 t_\alpha(aX^{-1}) (x_{-\alpha}(mX), 1, 1) = \left[ x_\alpha(aX^{-1})\cdot x_{-\alpha}(mX) \cdot x_\alpha\left(-\tfrac{aX^{-1}}{1 + am}\right) \{ X, 1+am \} ,\ x_\alpha\left(\tfrac{aX^{-1}}{1 + am}\right),\  (1 + am)\right] = \\
 = \left[ x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) ,\ x_\alpha\left(\tfrac{aX^{-1}}{1 + am}\right),\  (1 + am)\right]
\end{multline*}

This implies that 
\begin{multline*} x_{\alpha}\left(\tfrac{-aX^{-1}}{1+am}\right) \cdot h_\alpha(1+am)^{-1} \cdot \langle a, m \rangle^{-1} \cdot x_{-\alpha}\left(\tfrac{-mX}{1+am}\right) x_\alpha(aX^{-1}) \cdot x_{-\alpha}(mX) [1, 1, 1] = [1, 1, 1+am]
\end{multline*}

\begin{equation} \langle a,m \rangle _ \alpha = x_{-\alpha}\left(\tfrac{- m}{1 + am}\right) \cdot x_{\alpha}(a) \cdot x_{-\alpha}(m) \cdot x_{\alpha}\left(\tfrac{- a}{1+am}\right) \cdot h_{\alpha}^{-1}(1 + am). \end{equation} 
\begin{equation} \langle a,m \rangle _ \alpha ^{-1} =  h_{\alpha}(1 + am) \cdot x_{\alpha}\left(\tfrac{a}{1+am}\right) \cdot x_{-\alpha}(-m) \cdot x_{\alpha}(-a) \cdot x_{-\alpha}\left(\tfrac{m}{1 + am}\right). \end{equation} \end{comment}

\begin{externaltheorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{externaltheorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B^-)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g = i_-(\gamma)$ for some $\gamma \in G_M^0$ satisfying $j_+i_+(\gamma) = 1$,
  which implies that $g = 1$.

@@

 Set $h_+ = g_+g_0^{-1}$ and $h_- = g_-g_0^{-1}$.
 It is clear that \[ h_+ \in \overline{\St}(\Phi, A[X], M[X]),\ h_- \in \overline{\St}(\Phi, A[X\inv], M[X\inv])\text{ and that }j_-(h_-) = j_+(h_+).\]
 From~\cref{prop43} and~\cref{lem:action} we obtain that
 \[ [j_+(h_+)^{-1}, h_-, 1] = j_+(h_+)^{-1} \cdot [1, h_-, 1] = j_+(h_+)^{-1} j_B^-(h_-) \cdot [1, 1, 1] = [1, 1, 1]. \]
 Thus, $h_-$ (and hence $h_+$) come from some element of $G_M^0$, which implies that $g_+$ and $g_-$ are images of some element of $\St(\Phi, A)$.
 \end{proof}

\begin{proof}[Proof of~\cref{thm:main}]
 Notice that the assertions of the theorem for $\KO_2(2\ell, -)$ and $\K_2(\rD_\ell, -)$ follow from the assertion for $\St(\rD_\ell, -)$.
 In turn, the latter assertion follows from~\cref{thm41} and~\cref{thm:P1glueing} in the special case when $A$ is a local ring.
